---
title: "Reproducible Research: Peer Assessment 1"
author: "Ted M"
date: "October 23, 2016"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

```{r cache=TRUE}
# unzip and read the data from the "activity.csv" file, and parse the date field
unzip("repdata%2Fdata%2Factivity.zip")
activity <- read.csv("activity.csv")
activity$date <- as.Date(activity$date, format="%Y-%m-%d")

```

## What is mean total number of steps taken per day?

```{r}
# summarize the dataframe into daily total activity
suppressPackageStartupMessages(library(dplyr))
dailyActivity <- activity %>% group_by(date) %>% summarize(steps=sum(steps, na.rm=TRUE))
# plot the total steps per day as a histogram
plot(dailyActivity$date, dailyActivity$steps, type="h", lwd=4, main="Total steps per day", xlab="Date", ylab="Steps")
# functions to calculate the mean and median # of steps per day, ignoring NA's
meanstepsperday <- function() {mean(dailyActivity$steps, na.rm=TRUE)}
medianstepsperday <- function() { median(dailyActivity$steps, na.rm=TRUE)}
```
The mean number of steps per day is **`r meanstepsperday()`**  
The median number of steps per day is **`r medianstepsperday()`**

## What is the average daily activity pattern?

```{r}
# summarize the data across all days by 5-min interval
meanActivity <- activity %>% group_by(interval) %>% summarize(avgSteps=mean(steps, na.rm=TRUE))
plot(meanActivity$interval, meanActivity$avgSteps, type="l", lwd=4, main="Average steps per 5-min Interval", xlab="Interval#", ylab="Steps")
# function to find the interval# which has the highest average steps
maxAvgStepsInterval <- function() {meanActivity[meanActivity$avgSteps==max(meanActivity$avgSteps, na.rm = TRUE),]$interval}
```
The maximum average number of steps occurs in interval# **`r maxAvgStepsInterval()`**

## Imputing missing values

```{r}
# 1. Calculate and report the total number of missing values in the dataset
# function to count the number of steps observations that are NA
naCount <- function() {sum(is.na(activity$steps))}
# 2.Devise a strategy for filling in all of the missing values in the dataset
# derive "imputedSteps", by joining to the meanActivity dataset by interval#,
# then replace missing values with the mean for the interval.
# add "avgSteps" column to activity data
adjusted <- activity %>% left_join(meanActivity, by=c("interval"))
# set the imputedSteps column (integer) to be the first non-NA value of steps, round(avgSteps)
adjusted <- adjusted %>% mutate(imputedSteps = coalesce(steps, as.integer(round(avgSteps))))
# select and rename columns to look like the original dataframe
adjusted <- adjusted %>% select(steps=imputedSteps, date, interval)

#Make a histogram of the total number of steps taken each day and Calculate and report the mean
# and median total number of steps taken per day. Do these values differ from the estimates from
# the first part of the assignment? What is the impact of imputing missing data on the estimates
# of the total daily number of steps?
adjustedDailyActivity <- adjusted %>% group_by(date) %>% summarize(steps=sum(steps, na.rm=TRUE))
# plot the total steps per day as a histogram
plot(adjustedDailyActivity$date, adjustedDailyActivity$steps, type="h", lwd=4, main="Total steps per day", xlab="Date", ylab="Steps")
# functions to calculate the mean and median # of steps per day, ignoring NA's
adjustedmeanstepsperday <- function() {mean(adjustedDailyActivity$steps, na.rm=TRUE)}
adjustedmedianstepsperday <- function() { median(adjustedDailyActivity$steps, na.rm=TRUE)}
```
- The number of observations with missing value is **`r naCount()`**
- The adjusted mean number of steps per day is **`r adjustedmeanstepsperday()`**  
- The adjusted median number of steps per day is **`r adjustedmedianstepsperday()`**
- The impact of imputing missing data on the estimates is to: 
- a) increase the mean and median daily steps counts
- b) the Average Steps per 5-min interval remains mostly unchanged



## Are there differences in activity patterns between weekdays and weekends?
